---
title: "Assignment 3"
author: "Lauren"
date: "November 2, 2018"
output: html_document
---

##Q2. 
**I decided to add a small value(0.1) to all of my values before calculating the log2. I chose this option because it prevents me from having to remove the missing values before calculating the log2 and should give a more accurate mean. If I were to remove all zero values then the mean would not be accounting for these zeros, and thus would be skewed to a higher mean. In this case, it is important to note that adding small values to zero creates a log2 of negative value.** 
```{r}
#import the data to R
rna_counts <- read.csv("~/downloads/eXpress_dm_counts.csv", header=TRUE, row.names = 1, sep = ",")

#define the mean_expression function
mean_expression <- function(x, log=TRUE) { 
        if (log==FALSE) {
           mean(x) 
         } else { 
             z<-x+0.1 
             mean(log2(z)) 
       } 
}

#test the function
mean_expression(rna_counts[[1]])
mean_expression(rna_counts[[2]])
mean_expression(rna_counts[[3]])

mean_expression(rna_counts[[1]], log=FALSE)
mean_expression(rna_counts[[2]], log=FALSE)
mean_expression(rna_counts[[3]], log=FALSE)
```

##Q3. 
```{r}
#use a for loop to calculate the log transformed total_mean_expression of each sample
mean_cols <- vector(length=55)
for (i in 1:ncol(rna_counts)) {
    mean_cols[i] <- mean_expression(rna_counts[, i])
    col <- colnames(rna_counts)
    total_mean_expression_log <- cbind(col, mean_cols)
  }

head(total_mean_expression_log)

#use a for loop to calculate the total_mean_expression of each sample
for (i in 1:ncol(rna_counts)) {
    mean_cols[i] <- mean_expression(rna_counts[, i], log=FALSE)
    col <- colnames(rna_counts)
    total_mean_expression <- cbind(col, mean_cols)
  }

head(total_mean_expression)

#look for patterns in expression
total_mean_expression_matrix <- as.matrix(total_mean_expression)
total_mean_expression_matrix[order(total_mean_expression_matrix[mean_cols])]
```
**Males have higher expression levels than females.**

##Q4.  
```{r}
#use sapply to calculate the mean expression of each sample
total_mean_expression_sapply_log <- sapply(rna_counts, mean_expression)
total_mean_expression_sapply <- sapply(rna_counts, mean_expression, log=F)

head(total_mean_expression_sapply_log)
head(total_mean_expression_sapply)
```


```{r}
#calculate which method is faster 
system.time(for (i in 1:ncol(rna_counts)) {
  mean_cols[i]<-mean_expression(rna_counts[, i])
  col<-colnames(rna_counts)
  total_mean_expression <- cbind(col, mean_cols)
   })

system.time(total_mean_expression_sapply<-sapply(rna_counts, mean_expression))

```

  **The sapply method is faster.**

##Q5. 
```{r}
#use colMeans function to calculate mean expression of each sample
total_mean_expression_easy <- colMeans(rna_counts)
head(total_mean_expression_easy)

#or to get it with log2 transformed data
total_mean_expression_log_easy <- colMeans(log2(rna_counts+0.1))
head(total_mean_expression_log_easy)
```

##Q6.
```{r}
#calculate the mean expression of each gene
mean_gene_expression <- rowMeans(rna_counts)
head(mean_gene_expression)

#or to get it with log2 transformed:
mean_gene_expression_log <- rowMeans(log2(rna_counts+0.1))
head(mean_gene_expression_log)
```

##Q7.

```{r}
#calculate the mean gene expression for both large and small male head horn
combined_gene_expression <- rowMeans(rna_counts[ ,grep("M*_*_male_hdhorn", col, value=TRUE)])
head(combined_gene_expression)

#calculate the mean gene expression for small male head horn
sm_gene_expression <- rowMeans(rna_counts[ ,grep("M*_sm_male_hdhorn", col, value=TRUE)])
head(sm_gene_expression)

#calculate the mean gene expression for large male head horn
lg_gene_expression <- rowMeans(rna_counts[ ,grep("M*_lg_male_hdhorn", col, value=TRUE)])
head(lg_gene_expression)

#calculate the difference in mean gene expression between large and small male head horn
difference_mean_expression <- lg_gene_expression - sm_gene_expression
head(difference_mean_expression)
```

##Q8.

```{r}
#calculate the log transformed mean gene expression for large and small male head horn
combined_gene_expression_log <- log2(rowMeans(rna_counts[ ,grep("M*_*_male_hdhorn", col, value=TRUE)] + 0.1))

#calculate the log transformed mean gene expression for small male head horn
sm_gene_expression_log <- log2(rowMeans(rna_counts[ ,grep("M*_sm_male_hdhorn", col, value=TRUE)] + 0.1))

#calculate the log transformed mean gene expression for large male head horn
lg_gene_expression_log <- log2(rowMeans(rna_counts[ ,grep("M*_lg_male_hdhorn", col, value=TRUE)] + 0.1))

#calculate the log transformed mean gene expression difference between large and small male head horn
difference_mean_expression_log <- lg_gene_expression_log - sm_gene_expression_log

#graph the MAplot
library(ggplot2)
ggplot(rna_counts, aes(x=combined_gene_expression, y=difference_mean_expression))+
    geom_point()+
  ggtitle("Difference in Gene Expression Between Small and Large
          Male Asian Rhinocerous Beetle Head Horns")+
  labs(x="Combined Mean Gene Expression of Large and Small Males", y="Difference in Mean Gene Expression 
       Between Small and Large Males")

#graph the MAplot using log transformed data
ggplot(rna_counts, aes(x=combined_gene_expression_log, y=difference_mean_expression_log))+
    geom_point()+
  ggtitle("Difference in Gene Expression Between Small and Large
          Male Asian Rhinocerous Beetle Head Horns") +
  labs(x="Combined Mean Gene Expression of Large and Small Males", y="Difference in Mean Gene Expression 
       Between Small and Large Males (log2(mean))")
```

##Bonus.
```{r}
library(dplyr)
#Using the tidyverse, although not necessarily easier, another way to calculate the mean gene expression of each gene in small and large males
combined_gene_expression_tidy <- rna_counts %>%
     select(colnames(rna_counts[ , grep("M*_*_male_hdhorn", col)])) %>%
     rowMeans()
sm_gene_expression_tidy <- rna_counts %>%
     select(colnames(rna_counts[ , grep("M*_sm_male_hdhorn", col)])) %>%
     rowMeans()
lg_gene_expression_tidy <- rna_counts %>%
     select(colnames(rna_counts[ , grep("M*_lg_male_hdhorn", col)])) %>%
     rowMeans()
difference_mean_expression_tidy <- lg_gene_expression_tidy - sm_gene_expression_tidy

#graph the result
ggplot(rna_counts, aes(x=combined_gene_expression_tidy, y=difference_mean_expression_tidy))+
    geom_point()+
  ggtitle("Difference in Gene Expression Between Small and Large
          Male Asian Rhinocerous Beetle Head Horns")+
  labs(x="Combined Mean Gene Expression of Large and Small Males", y="Difference in Mean Gene Expression 
       Between Small and Large Males")

#for log transformed mean gene expression
combined_gene_expression_tidy_add <- rna_counts %>%
     select(colnames(rna_counts[ , grep("M*_*_male_hdhorn", col)])) + 0.1

combined_gene_expression_tidy_log <- log2(rowMeans(combined_gene_expression_tidy_add))

sm_gene_expression_tidy_add <- rna_counts %>%
     select(colnames(rna_counts[ , grep("M*_sm_male_hdhorn", col)])) + 0.1
     
sm_gene_expression_tidy_log <- log2(rowMeans(sm_gene_expression_tidy_add))

lg_gene_expression_tidy_add <- rna_counts %>%
     select(colnames(rna_counts[ , grep("M*_lg_male_hdhorn", col)])) + 0.1
     
lg_gene_expression_tidy_log <- log2(rowMeans(lg_gene_expression_tidy_add))

difference_mean_expression_tidy_log <- lg_gene_expression_tidy_log - sm_gene_expression_tidy_log

#graph the result
ggplot(rna_counts, aes(x=combined_gene_expression_tidy_log, y=difference_mean_expression_tidy_log))+
    geom_point()+
  ggtitle("Difference in Gene Expression Between Small and Large
          Male Asian Rhinocerous Beetle Head Horns")+
  labs(x="Combined Mean Gene Expression of Large and Small Males", y="Difference in Mean Gene Expression 
       Between Small and Large Males")

```








