---
title: "Week7"
author: "Lauren"
date: "October 29, 2018"
output: html_document
---

---
title: "In class assignment week 2, part 2. A worked example using control flow (for loops, if statements, etc)"
author: "Ian Dworkin"
output: 
  html_document:
    keep_md: yes
    number_sections: yes
---

# Introduction
Let's do a little exercise integrating some of the things we have learned. Here are some Illumina HiSeq reads for one of our recent projects:

```{r}
read_1 <- "CGCGCAGTAGGGCACATGCCAGGTGTCCGCCACTTGGTGGGCACACAGCCGATGACGAACGGGCTCCTTGACTATAATCTGACCCGTTTGCGTTTGGGTGACCAGGGAGAACTGGTGCTCCTGC"

read_2 <- "AAAAAGCCAACCGAGAAATCCGCCAAGCCTGGCGACAAGAAGCCAGAGCAGAAGAAGACTGCTGCGGCTCCCGCTGCCGGCAAGAAGGAGGCTGCTCCCTCGGCTGCCAAGCCAGCTGCCGCTG"

read_3  <- "CAGCACGGACTGGGGCTTCTTGCCGGCGAGGACCTTCTTCTTGGCATCCTTGCTCTTGGCCTTGGCGGCCGCGGTCGTCTTTACGGCCGCGGGCTTCTTGGCAGCAGCACCGGCGGTCGCTGGC"
```

Question 1. what species are these sequences from?

Drosophila melanogastor

Question 2. Put all three of these reads into a single object (a vector).  What class will the vector `reads` be? Check to make sure! How many characters are in each read (and why does `length()` not give you what you want.. try...)

> reads<-c(read_1, read_2, read_3)
The class is character. Length() gives an answer of 3. There are 124 characters in each read(nchar(reads)).

Question 3. Say we wanted to print each character (not the full string) from read_1, how do we do this using a for loop? You may wish to look at a function like `strsplit()` to accomplish this (there are other ways.)

Replace the blanks.
```{r}
read_1_split <- strsplit(read_1, split = character(0), fixed = T) 
```

Using a for loop:

> for (x in read_1) {
+     print(strsplit(read_1, split=character(0), fixed=TRUE))
+ }

Question 4. What kind of object does this return? How might we make it a character vector again?

This gives a list (class(read_1_split)). To convert it back to a character string use:
read_1_split_new<-as.character(unlist(read_1_split))

Question 5. How about if we wanted the number of occurrences of each base? Or better yet, their frequencies? You could write a loop, but I suggest looking at the help for the `table()` function... Also keep in mind that for for most objects `length()` tells you how many elements there are in a vector. For lists use `lengths()` (so you can either do this on a character vector or a list, your choice)

To calculate the number of each bases use: table(read_1_split_new)
To calculate the frequencies: use table(read_1_split_new)/length(read_1_split_new)

Question 6. How would you make this into a nice looking function that can work on either  a list or vectors of characters? (Still just for a single read)

BaseFrequencies<_ function(x) {
if (mode(x) == "list") {
tab <- table(x)/lengths(x)
} else{
tab<- table(x)/lengths(x)
}
}

Question 7. Now how can you modify your approach to do it for an arbitrary numbers of reads? You could use a loop or use one of the apply like functions (which one)?

if (length(x) == 1 & mode(x) == "character") {
x<- strsplit(x, split = " ", fixed=T)
x<- as.character(unlist(x))
}
if (mode(x) == "list")
tab<- table(x)/lengths(x)
else (tab <- table(x)/length(x))


Question 8. Can you revise your function so that it can handle the input of either a string as a single multicharacter vector, **or** a vector of individual characters **or** a list? Try it out with the vector of three sequence reads (`reads`).  

if (length(x) == 1 & mode(x) == "character") {
x<- strsplit(x, split = " ", fixed=T)
x<- as.character(unlist(x))
}
if (mode(x) == "list") {
tab<- table(x)/lengths(x)
} else { (tab <- table(x)/length(x))
}
BaseFreq<-sapply(reads, BaseFrequencies, USE.NAMES=F)
print(BaseFreq, digits=2)

title: "Bio720 R In Class Exercise 2"
author: "Ian Dworkin"
date: "October 29, 2018"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
---
# Introduction

In class assignment. Please work in groups of 2-3. When you are finished with your assignment, please (as both a .Rmd and a .md file) add this to your class github repo (or make a new one if you really want). 

## for loops and the apply family of functions
During the past two week for, you learned a bit about how to use `for` loops and `apply` family of functions. Now we are going to do some in class practice and extend your understanding of how they work. In case you want some alternative tutorials [here](https://www.datacamp.com/community/tutorials/tutorial-on-loops-in-r#gs.FgFVHlY) is one that goes into more detail.

1. What happens to `x` for each of these functions after you run them? Please explain what is happening in each case. i.e. Why do these behave differently?

```{r}
x <- 1

for (i in 1:9) {
    x <- x + 1
    print(x)
}

x

```

**VS**

```{r, error=TRUE}
rm(x)

countFun <-function(x)
{

  for (i in 1:10) {
    x <- x + 1
    print(x)
}}

countFun(1)
x
```

'x' in the first example is assigned outside of the function (in the global environment) so it is always 1. In the second example x is part of the function and does not exist outside of that function.

2. Write a for loop that generates a vector which is the square of all integers from 1 to 10000. Make this all happen within the loop. i.e. We are going to use the `system.time()` function to monitor how long it takes. This is pretty coarse, and a way better way of doing this is with the `microbenchmark` package (which is worth learning), but for now this function will do.

system.time(for (i in 1:10000) {
print(i^2)
})
  user  system elapsed 
  0.769   0.045   1.392 

3.  Now do the same thing as in question 2, but *pre-allocate* memory for the vector `x` (or whatever you want to call it) to store the 10000 values *before* running the `for` loop. You don't need to do anything differently, just initialize a vector of length 10000. Which runs faster? Please explain why?

x<- 1:10000
system.time(for (i in x) {
print(i^2)
})

   user  system elapsed 
  0.765   0.047   1.312 

It is faster because it doesn't have to generate the range that is already defined in vector x.

4. What is a very "`R`" to make this (question 2 and 3) even faster? (i.e. think about vectorization)

Use:
x_sq<- system.time(for (i in x) {
    print(i^2)
})
print(x_sq)

5. Why is this faster? So what is the lesson learned from this?

This stores the result in a new vector so R does not have to print the result while it's running the loop.

6 - Compare the following pieces of code for both what they are doing, and for speed (you can wrap this in `system.time()`). We are using the `rnorm` function which generates random numbers from a normal distribution (with a particular mean and standard deviation)

```{r}
n <- 10000 # just run this once, each time you change n

rm(x)
system.time(x <- replicate(n = n, expr = rnorm(1, mean = 5, sd = 5)))

```

**VS**

```{r}
rm(x)
x <- rep(NA, n)
system.time(for (i in 1:n) x[i] <- rnorm(1, mean = 5, sd = 5))
```

Repeat each of these 5 times, and write down how long they take. Now repeat this for `n = 100000` iterations. Repeat for `n = 1000000`

6. (continued) Is there a difference in speed? Can you explain why? What are these two pieces of code doing differently?

The second option is faster. This is because x is already defined outside of the for loop. When n is increased, it takes longer because there is more numbers to work with.

7. Can you think of a faster way to generate the random numbers? hint, look at the help for `rnorm`.

## Running a power analysis.

In statistics, we often want to check (before doing an experiment) how likely we will be able to detect "an effect". This is true whether we are doing RNAseq, or a simple t-test. A tests statistical power is a function of three things. First, the magnitude of the observed effect (i.e. does your experiment have a big effect on gene expression or really small subtle effects). Second the amount of variation within each treatment group (if what you measure is really noisy, it will make it more difficult to get accurate estimates) and of course sample size. So when statistically oriented researchers are testing new methods, or just planning experiments, they will often use simulations to perform power analyses.  These require repeating the same simulations over and over again and checking results against some expectation or some hypothetical threshold (i.e. statistical significance is something some researchers care about... although you should focus on magnitude of effect size and confidence intervals... but that is another conversation).

Here is an example of a very simple simulation, which generates a linear relationship between two variables (the dependent variable *x* and the response variable *y*), simulates values of *x* based on this relationship and the amount of variation (noise) and then fits a linear regression and extracts the p-value associated with the slope of the relationship.

```{r}
rm(list=ls())

# sample size, n
n <- 20

# intercept, a
a = 3

# slope, b
b = 0.3

# independent/explanatory variable
x <- rnorm( n = n, mean = 10, sd = 2)

# response/dependent variable
y <- rnorm(n = length(x), mean = a + b*x, sd = 1 )

plot(y ~ x, pch = 20, cex = 1.5)

# use the `lm` function to fit a linear model (including   a regression like here)
mod_1 <- lm(y ~ x)

summary(mod_1) # just to look at.

(p_val <- summary(mod_1)$coef[2,4])

```

Run the code above a few times just to convince yourself that you are generating "new data" each time, and as such the results from the regression are different (including the estimated slope, and associated stats like p-values.)

8. How would you take this idea and create a vector that stores the p_values from these simulations, and allows you to repeat the simulation 1000 times using a `for` loop? Then use this vector to find out what proportion of p-values are less than 0.05. You can finish this code up by using a histogram (with the `hist` function) to look at the distribution of p-values.


9. How would you do the same thing as in question 9 but by writing an explicit function and using `replicate` instead of a `for` loop? Why are the results from question 8 and 9 not the same?


You can use this to get more involved. Say you wanted to see what would happened if you varied the sample size from n = 10 to n = 100 or varied the slope from 0 to 0.5? How would you do this using both approaches from questions 9 and 10 (you may need to use nested for loops). I will show you in the answers.


